### 2016-10-02 paper/1_introduction.tex
Minor fixes

%In order to develop Big Data solutions, one has to solve multifaceted problems. %The necessary technology stack comprises the recording of data, the data cleaning, the meta-data generation, the representation and integration with other data sources, the analysis and modeling of the actual problem as well as the interpretation. Each task for itself is highly complex and deserves an individual paper.
%For us, the data analysis part is the key aspect in solving Big Data problems. Henceforth, we will concentrate on how to gain valuable insights from a huge data set. Once big data sets were amassed, people quickly recognized that these sets contain valuable information they only have to harness. 
Key component of modern big data solutions are a sophisticated analysis algorithms.% In this work, we focus on the data analysis part of the problems. 
Unfortunately, the search for useful patterns and peculiarities in large data sets resembled the search for a needle in a haystack. This challenge requires tools that are able to analyze vast amounts of data quickly. Most of these tools are based on statistics to extract interesting features. It is beneficial to apply these statistical means to the complete data set instead of smaller chunks. Even if the chunks cover the complete data set, important correlations between data points might get lost if chunks are treated individually. Moreover, the statistical tools improve their descriptive and predictive results by getting fed more data. The more data is available, the more likely it is that noise cancels out and that the significant patterns manifest. However, these benefits come at the price of an increased computing time, which requires fast computer systems to make computations feasible. 
%The insights gained from collected data already help to govern business decisions, improve life quality or simply create new industries from scratch.
%For example, retail stores analyze their sales, customer, pricing and weather data in order to decide which products to offer or when to do a discount sale. That not only increases the revenue of the shops but also boosts the satisfaction of the customers by getting better offers. 
%Police departments try to detect probable crime sites by extracting patterns from previously recorded criminal acts and then reinforce the policemen in that region~\cite{lohr:yt2012a}.
%The intelligent deployment of policemen improves security for citizens without having to hire more officers.
%Hospitals analyze their patients' records and scientific studies in order to find the cancer treatment best complying with the specifics of the patient. The individual therapy maximizes the chance of cure~\cite{watson:2013a}.
%Another example showing the benefits of data analysis made it even into a Hollywood film.
%The film Moneyball is based on the true story of the Oakland Athletics baseball team, whose management built an elite team in an unfavorable financial situation. They employed sophisticated data analysis methods to spot underestimated players, who they could cheaply recruit.
%That marked the start of sports statistics, which are nowadays a common tool for professional teams.
%These examples emphasize the importance and utility of information gained with analytic tools from gathered data.
%Interestingly, the IDC, an international market research firm, estimates that only $0.5\%$ of the globally collected data is harnessed~\cite{gantz:iaf2012a}. They further state that about $23\%$ of today's data is worth being analyzed. Thus, there is still potential left for improvements.

@@

By finding an algorithm with a lower runtime complexity than an existing algorithm, e.g., for clustering, run times can be drastically decreased. 

@@

With this new models, programmers can concentrate on the actual algorithm and use case rather than reasoning about distributed systems. These are the reasons why Google's MapReduce~\cite{dean:c2008a} framework and its open source re-implementation Hadoop~\cite{hadoop:2008a} became so popular among scientists as well as engineers.

@@

\begin{figure}[t!]
\centering
\includegraphics[height=0.175\paperheight]{images/systemArchitecture.png}
\caption{The layered system architecture of Gilbert. The language layer is responsible for parsing, typing and compiling the given Matlab code. The intermediate layer facilitates high-level optimization strategies. The runtime layer is responsible for executing the specified program in parallel.}
\label{fig:systemArchitecture}
\end{figure}

As a solution we propose Gilbert, a distributed sparse linear algebra environment. Gilbert provides a Matlab-like language for distributed sparse linear algebra operations. It has a layered architecture, which is shown in Figure \ref{fig:systemArchitecture}. The first layer is the language layer. It contains all functionality to parse the given Gilbert code and to compile it into an intermediate representation (IR). The second layer is the intermediate layer and it receives the IR of the Gilbert code. The intermediate format is the ideal representation to apply language independent high-level transformations. The Gilbert optimizer applies several algebraic optimizations prior to the generation of the execution plan. The execution plan generator translates the optimized IR into a specific execution plan, depending on the selected execution engine. Once the program has been translated into an execution engine's specific plan, it is executed on the respective back end.

Our main contributions are:
\begin{itemize}
 \item Gilbert allows to write MATLAB-like code for sparse linear algebra and execute it on massively parallel dataflow systems.
 \item Its expressiveness allows to quickly develop scalable algorithms to analyze web-scale data.
 \item We introduce a novel fixed-point operator which replaces loops by recursion and can be efficiently parallelized.
 \item Furthermore, we demonstrate that dataflow optimizers can be used to automatically select an good matrix multiplication strategy. 
\end{itemize}
%Gilbert provides a Matlab-like language for distributed sparse linear algebra operations. %Being such a system, it comprises the complete stack of functionalities necessary to implement a programming language. At first, the system has to divide the given source code into tokens. These tokens are parsed and an abstract syntax tree (AST) is created. In order to generate the intermediate representation (IR), type information is needed for all expressions. This information is extracted by the typing system. Knowing the types, the compiler can translate the Gilbert program into its IR. The IR is well suited to apply high-level transformations to the program. At last, the execution plan generator translates the IR into an execution plan which can be executed in parallel. %The layered system architecture is shown in \cref{fig:systemArchitecture}.

>>>>

%In order to develop Big Data solutions, one has to solve multifaceted problems. %The necessary technology stack comprises the recording of data, the data cleaning, the meta-data generation, the representation and integration with other data sources, the analysis and modeling of the actual problem as well as the interpretation. Each task for itself is highly complex and deserves an individual paper.
%For us, the data analysis part is the key aspect in solving Big Data problems. Henceforth, we will concentrate on how to gain valuable insights from a huge data set. Once big data sets were amassed, people quickly recognized that these sets contain valuable information they only have to harness. 
Key component of modern big data solutions are sophisticated analysis algorithms.% In this work, we focus on the data analysis part of the problems. 
Unfortunately, the search for useful patterns and peculiarities in large data sets resembles the search for a needle in a haystack. 
This challenge requires tools that are able to analyze vast amounts of data quickly. 
Most of these tools are based on statistics to extract interesting features. 
It is beneficial to apply these statistical means to the complete data set instead of smaller chunks. 
Even if the chunks cover the complete data set, important correlations between data points might get lost if chunks are treated individually. 
Moreover, the statistical tools improve their descriptive and predictive results by getting fed more data. 
The more data is available, the more likely it is that noise cancels out and that significant patterns manifest. 
However, these benefits come at the price of an increased computing time, which requires fast computer systems to make computations feasible. 
%The insights gained from collected data already help to govern business decisions, improve life quality or simply create new industries from scratch.
%For example, retail stores analyze their sales, customer, pricing and weather data in order to decide which products to offer or when to do a discount sale. That not only increases the revenue of the shops but also boosts the satisfaction of the customers by getting better offers. 
%Police departments try to detect probable crime sites by extracting patterns from previously recorded criminal acts and then reinforce the policemen in that region~\cite{lohr:yt2012a}.
%The intelligent deployment of policemen improves security for citizens without having to hire more officers.
%Hospitals analyze their patients' records and scientific studies in order to find the cancer treatment best complying with the specifics of the patient. The individual therapy maximizes the chance of cure~\cite{watson:2013a}.
%Another example showing the benefits of data analysis made it even into a Hollywood film.
%The film Moneyball is based on the true story of the Oakland Athletics baseball team, whose management built an elite team in an unfavorable financial situation. They employed sophisticated data analysis methods to spot underestimated players, who they could cheaply recruit.
%That marked the start of sports statistics, which are nowadays a common tool for professional teams.
%These examples emphasize the importance and utility of information gained with analytic tools from gathered data.
%Interestingly, the IDC, an international market research firm, estimates that only $0.5\%$ of the globally collected data is harnessed~\cite{gantz:iaf2012a}. They further state that about $23\%$ of today's data is worth being analyzed. Thus, there is still potential left for improvements.

@@

By finding an algorithm with a lower runtime complexity than an existing algorithm, e.g. for clustering, run times can be drastically decreased. 

@@

With these new models, programmers can concentrate on the actual algorithm and use case rather than reasoning about distributed systems. These are the reasons why Google's MapReduce~\cite{dean:c2008a} framework and its open source re-implementation Hadoop~\cite{hadoop:2008a} became so popular among scientists as well as engineers.

@@

\begin{figure}[t!]
\centering
\includegraphics[height=0.175\paperheight]{images/systemArchitecture.png}
\caption{The layered system architecture of Gilbert. The language layer is responsible for parsing, typing and compiling the given MATLAB code. The intermediate layer facilitates high-level optimization strategies. The runtime layer is responsible for executing the specified program in parallel.}
\label{fig:systemArchitecture}
\end{figure}

As a solution we propose Gilbert, a distributed sparse linear algebra environment. Gilbert provides a MATLAB-like language for distributed sparse linear algebra operations. It has a layered architecture, which is shown in Figure \ref{fig:systemArchitecture}. The first layer is the language layer. It contains all functionality to parse the given Gilbert code and to compile it into an intermediate representation (IR). The second layer is the intermediate layer and it receives the IR of the Gilbert code. The intermediate format is the ideal representation to apply language independent high-level transformations. The Gilbert optimizer applies several algebraic optimizations prior to the generation of the execution plan. The execution plan generator translates the optimized IR into a specific execution plan, depending on the selected execution engine. Once the program has been translated into an execution engine's specific plan, it is executed on the respective back end.

Our main contributions are:
\begin{itemize}
 \item Gilbert allows to write MATLAB-like code for sparse linear algebra and execute it on massively parallel dataflow systems.
 \item Its expressiveness allows to quickly develop scalable algorithms to analyze web-scale data.
 \item We introduce a novel fixed-point operator which replaces loops by recursion and can be efficiently parallelized.
 \item Furthermore, we demonstrate that dataflow optimizers can be used to automatically select a good matrix multiplication strategy. 
\end{itemize}
%Gilbert provides a Matlab-like language for distributed sparse linear algebra operations. %Being such a system, it comprises the complete stack of functionalities necessary to implement a programming language. At first, the system has to divide the given source code into tokens. These tokens are parsed and an abstract syntax tree (AST) is created. In order to generate the intermediate representation (IR), type information is needed for all expressions. This information is extracted by the typing system. Knowing the types, the compiler can translate the Gilbert program into its IR. The IR is well suited to apply high-level transformations to the program. At last, the execution plan generator translates the IR into an execution plan which can be executed in parallel. %The layered system architecture is shown in \cref{fig:systemArchitecture}.

### 2016-10-02 paper/3_gilbert_features.tex
Minor fixes

Gilbert's language has been strongly inspired by MATLAB, which should ease its adoption.
The language is a fully functional subset of MATLAB and its grammar is specified in \cite{Rohrmann2014}.
The elementary data type is the matrix. Gilbert supports arbitrary $2$-dimensional matrices whose elements can be \emph{double} or \emph{boolean}.
Vectors are not represented by a special type but instead are treated as a single column/row matrix.
Additionally, scalar \emph{double}, \emph{boolean}, and string values are supported. Gilbert also implements cell array types.
A cell array is defined using curly braces and commas to separate individual values. The cells can be accessed by an index appended in curly braces to the cell array variable. \Cref{lst:cellArray} shows how to define and access them.

@@

\begin{listing}
  \centering
  \begin{sublisting}{.4\linewidth}
    \begin{lstlisting}[language=Matlab,
      commentstyle=\color{black},
      stringstyle=\color{black},
      keywordstyle=\color{black}\bfseries,
    ]
A = 0;
for i = 1:10
  A = A + i;
end
    \end{lstlisting}
    \caption{For loop}
    \label[listing]{fig:for2Fixpoint:for}
  \end{sublisting}
  \begin{sublisting}{.5\linewidth}
      \begin{lstlisting}[language=Matlab]
f = @(x) ...
  {x{1} + x{2}, x{2} + 1};
r = fixpoint({0,1}, f, 10);
A = r{1};
    \end{lstlisting}
    \caption{Fixpoint}
    \label[listing]{fig:for2Fixpoint:fixpoint}
  \end{sublisting}
  \caption{Transformation from Matlab for loop \subref{fig:for2Fixpoint:for} to Gilbert fixpoint \subref{fig:for2Fixpoint:fixpoint} formulation. Essentially, all iteration data is combined and passed as a cell array value to the update function.}
  \label{fig:for2Fixpoint}
\end{listing}

@@

The matrix size inference is incorporated into the HM type inference by adding some logic to the \code{unify} function.
Whenever we encounter a matrix type during the unification process, we call a \code{unifyValue} function on the two row and column values.
The \code{unifyValue} function works similarly to the \code{unify} function.
First, the function resolves the value expression, thus substituting value variables with their assigned values.
Then, if at least one of the resolved value expressions is still a value variable, then the union is constructed and the corresponding value variable dictionary entry is updated.
If both resolved expressions are equal, then this value is returned and otherwise a value mismatch error is thrown.

@@

\textbf{Transpose Pushdown}. Transpose pushdown tries to move the transpose operations as close to the matrix input as possible.
Thereby, consecutive transpose operations accumulate at the inputs and unnecessary operations erase themselves.
An example is given in \cref{lst:pushdown}. By inserting $C$ into $E$, the expression $E=(A^T BD^T)^T$ is obtained. This term is equivalent to $DB^T A$. The latter formulation contains only one transpose operation. Usually multiple transpose operations occur because they are written for convenience reasons at various positions in the code. Moreover, in complex programs it is possible that the programmer loses track of them or simply is unaware of the optimization potential. Therefore, transpose pushdown can be a beneficial optimization.

>>>>

Gilbert's language has been strongly inspired by MATLAB, which should ease its adoption.
The language is a fully functional subset of MATLAB and its grammar is specified in \cite{Rohrmann2014}.
The elementary data type is the matrix. Gilbert supports arbitrary $2$-dimensional matrices whose elements can be \emph{double} or \emph{boolean}.
Vectors are not represented by a special type but instead are treated as a single column/row matrix.
Additionally, scalar \emph{double}, \emph{boolean}, and \emph{string} values are supported. Gilbert also implements cell array types.
A cell array is defined using curly braces and commas to separate individual values. The cells can be accessed by an index appended in curly braces to the cell array variable. \Cref{lst:cellArray} shows how to define and access them.

@@

\begin{listing}
  \centering
  \begin{sublisting}{.4\linewidth}
    \begin{lstlisting}[language=Matlab,
      commentstyle=\color{black},
      stringstyle=\color{black},
      keywordstyle=\color{black}\bfseries,
    ]
A = 0;
for i = 1:10
  A = A + i;
end
    \end{lstlisting}
    \caption{For loop}
    \label[listing]{fig:for2Fixpoint:for}
  \end{sublisting}
  \begin{sublisting}{.5\linewidth}
      \begin{lstlisting}[language=Matlab]
f = @(x) ...
  {x{1} + x{2}, x{2} + 1};
r = fixpoint({0,1}, f, 10);
A = r{1};
    \end{lstlisting}
    \caption{Fixpoint}
    \label[listing]{fig:for2Fixpoint:fixpoint}
  \end{sublisting}
  \caption{Transformation from MATLAB for loop \subref{fig:for2Fixpoint:for} to Gilbert fixpoint \subref{fig:for2Fixpoint:fixpoint} formulation. Essentially, all iteration data is combined and passed as a cell array value to the update function.}
  \label{fig:for2Fixpoint}
\end{listing}

@@

The matrix size inference is incorporated into the HM type inference by adapting the \code{unify} function.
Whenever we encounter a matrix type during the unification process, we call a \code{unifyValue} function on the two row and column values.
The \code{unifyValue} function works similarly to the \code{unify} function.
First, the function resolves the value expression, thus substituting value variables with their assigned values.
Then, if at least one of the resolved value expressions is still a value variable, then the union is constructed and the corresponding value variable dictionary entry is updated.
If both resolved expressions are equal, then this value is returned and otherwise an error is thrown.

@@

\textbf{Transpose Pushdown}. Transpose pushdown tries to move the transpose operations as close to the matrix input as possible.
Thereby, consecutive transpose operations accumulate at the inputs and unnecessary operations erase themselves.
An example is given in \cref{lst:pushdown}. By inserting $C$ into $E$, the expression $E=(A^T BD^T)^T$ is obtained. This term is equivalent to $DB^T A$. The latter formulation contains only one transpose operation. Usually multiple transpose operations occur because they are written for convenience reasons at various places in the code. Moreover, in complex programs it is possible that the programmer loses track of them or simply is unaware of the optimization potential. Therefore, transpose pushdown can be a beneficial optimization.

### 2016-10-02 paper/4_gilbert_runtime.tex
Minor fixes

%!TEX root=main.tex
\section{Gilbert Runtime}
\label{sec:gilbertRuntime}

The Gilbert runtime is responsible for executing the compiled MATLAB code on a particular platform. For this purpose, it receives the intermediate representation of the code and translates it into the execution engine's specific format. Currently, Gilbert supports three different execution engines: \emph{ReferenceExecutor}, \emph{FlinkExecutor} and \emph{SparkExecutor}. The \emph{ReferenceExecutor} executes the compiled MATLAB code locally. For the distributed execution Gilbert supports the Spark and the Flink system as backends. The \emph{ReferenceExecutor} is an interpreter for the intermediate code.
It takes the dependency tree of a MATLAB program and executes it by evaluating the operators bottom-up. In order to evaluate an operator, the system first evaluates all inputs of an operator and then executes the actual operator logic. Since the program is executed locally, the complete data of each operator is always accessible and, thus, no communication logic is required. All input and output files are directly written to the local hard drive. In contrast to the \emph{ReferenceExecutor}, the \emph{FlinkExecutor} executes the program in parallel. It takes the dependency tree of a MATLAB program and translates it into a PACT plan. After the plan is generated, it is issued to the Flink system for parallel execution. This approach implies that the program is not directly executed by the executor. Instead, the executor represents just another translation step. The PACT plans are executed in parallel. Consequently, data structures are needed which can be distributed across several nodes and represent the commonly used linear algebra abstractions, such as vectors and matrices. Moreover, the linear algebra operations have to be adjusted so that they keep working in a distributed environment. Fortunately, Flink offers a rich and expressive API to easily realize distributed computations. The details of the distributed data structures and operations are explained in \cref{sec:DistributedMatrixRepresentation} and \cref{sec:LinearAlgebraOperations}. The \emph{SparkExecutor} is the second executor for distributed computations. In contrast to the \emph{FlinkExecutor}, it executes the MATLAB programs on top of the Spark system. Since Spark and Flink offer a similar set of programming primitives, they can both operate on the same data structures. Furthermore, even most of the linear algebra operations can be implemented similarly.

@@

One aspect of writing distributed algorithms is how the relevant data is distributed across several worker nodes. Since the distribution pattern directly influences the algorithms, one cannot consider them independently from one another. In Gilbert's use case, the main data structure are matrices. Thus, a partitioning for matrices has to be conceived which allows efficient algorithms working on the distributed data. Looking at a matrix, one easily finds a multitude of different partition schemes. A first idea could be to partition a matrix according to their rows or columns, as it is depicted in \cref{fig:rowPartitioning} and \cref{fig:columnPartitioning}. This scheme allows to represent a matrix as a set of vectors which is stored in a distributed fashion.
Furthermore, it allows an efficient realization of cellwise operations, such as $+,-,/$ or $.*$. In order to calculate the result of such an operation, we only have to join the corresponding rows of both operands and execute the operation locally for each pair of rows.

@@

However, this approach unveils its drawbacks when multiplying two equally partitioned matrices $A$ and $B$. In such a case, the row $r$ of $A$ and the complete matrix $B$ is needed to calculate the resulting row with index $r$. This circumstance implies a complete repartitioning of $B$. The repartitioning is especially grave, since $B$ has to be broadcasted to every row of $A$. In order to quantify the repartitioning costs of such a matrix multiplication, a simple cost model is developed. First of all, it is limited to modeling the communication costs, since network I/O usually constitutes the bottleneck of distributed systems and is thus the dominating factor. For the sake of simplicity, the multiplication of two quadratic matrices $A \in \mathbb{R}^{n\times n}$ and $B \in \mathbb{R}^{n\times n}$ giving the matrix $C\in \mathbb{R}^{n \times n}$ is considered. Moreover, we assume that there are $n$ worker nodes available, each of which receiving a single row of $A$ and $B$, respectively. Thus, $A$ and $B$ are row-wise partitioned. We further assume that rows with the same index are kept on the same worker node. Each row $a_r$ requires the complete knowledge of matrix $B$ in order to produce the row $c_r$. Therefore, every row $b_r$ has to be sent to all other worker nodes. Thus, the number of rows sent by each worker node is $n-1$. All of the $n$ worker nodes have to do the same. Consequently, the total number of sent messages is $n(n-1)$ and each message has a size of $n\alpha$ where $\alpha$ is the size of a matrix entry. Usually, each sending operation causes some constant overhead inflicted by resource allocation. Before sending the actual data over the network, memory to transfer the data to the network interface has to be allocated, network resources have to be reserved and a network connection with the remote peer has to be established. This overhead is denoted by $\Delta$. Since all sending operations occur in parallel, the costs caused by constant overhead are $(n-1)\Delta$. The total amount of data, which has to sent over the network, is $n^2(n-1)\alpha$. The network interconnection is assumed to guarantee every node a bandwidth $\nu$. Therefore, the time needed for sending the data is $\frac{n^2(n-1)\alpha}{\nu}$. These considerations lead to the following repartitioning cost model:

@@

Row and column partitioning are extreme forms of blocking. A less extreme form would be to split the matrix into equally sized quadratic blocks as shown in \cref{fig:quadraticBlockPartitioning}. In order to identify the individual blocks, each of them will be assigned a block row and block column index assigned. Thus, blocking adds some memory overhead in the form of index information. The blocks are distributed across the worker nodes. The block size directly controls the granularity of the partitioning. Increasing the block size will reduce the memory overhead of distributed matrices while reducing the degree of parallelism. Thus, the user has to adjust the block size value depending on the matrix sizes and the number of available worker nodes in order to obtain best performance. The quadratic block partitioning has similar properties like the row- and column-wise partitioning scheme when it comes to cellwise operations. We simply have to join corresponding blocks with respect to the pair of block row and column index and execute the operation on matching blocks locally. But how does this pattern performs for matrix multiplications? The assumptions are the same as before and additionally it is assumed that $n$ is a square number. Since the matrices $A$, $B$ and $C$ are equally partitioned into square blocks, indices will henceforth reference the block and not the matrix entry. In order to calculate the block $c_{ij}$, we have to know the block row $a_{i}$ and the block column $b_{j}$. The resulting block will be stored on the node $n_{ij}$ which already contains the blocks $a_{ij}$ and $b_{ij}$. Thus, each node $n_{ij}$ has to receive the missing $2\left(\sqrt{n}-1\right)$ blocks from the block row $a_{i}$ and block column $b_{j}$. In total, all worker nodes have to send $2n\left(\sqrt{n}-1\right)$ blocks. Each block has the size $n\alpha$. The total communication costs comprises the transmission costs and the network overhead:

@@

In contrast to RMM, CPMM calculates the outer products between $A^k$ and $B_k$ for all $k$. A mapper can group the $A^k$ and $B_k$ together so that a reducer can compute the outer products. Consequently, this method does not replicate any data. The outer product produces intermediate result matrices $C^k$ which have to be added up to produce the final result $C=\sum_{k=1}^{l} C^k$. This summation can be achieved by a subsequent reducer. The whole process is illustrated in \cref{fig:CPMM}. The methods RMM and CPMM differ in terms of network communication. The former method can be realized within a single MapReduce job whereas the latter requires two. Neither RMM nor CPMM is always superior. The optimal matrix multiplication strategy depends on the matrix size of its operands $A$ and $B$. Fortunately, Flink and Spark exhibit a little bit more flexibility in terms of higher order functions. Looking at the definition of the matrix multiplication for $C_{ij}=\sum_{k=1}^{l}A_{ik}\times B_{kj}$, it can be seen that every $A_{ik}$ has to be joined with its corresponding $B_{kj},\forall k$. This pairwise mapping can be easily achieved by using the join function. The join-key is the column index of $A$ and the row index of $B$. The joiner calculates for each matching pair $A_{ik}$ and $B_{kj}$ an intermediate result $C_{ij}^k$. Grouping the intermediate results with respect to the index pair $(i,j)$ allows us to compute the final result in a subsequent reduce step. The overall algorithm strongly resembles the CPMM. Flink supports several execution strategies for the higher-order functions. A cost-based optimizer selects the best strategies prior to execution. One possible optimization concerns the join function. The join can either be realized using a hybrid-hash join or a sort-merge join algorithm depending on the current partitioning and the input data sizes. If one input data is relatively small compared to the other input, it is usually more efficient to use the hybrid-hash join algorithm. Without loss of generality, we assume that the matrix $B$ constitutes such a small input. If we further assume that the block rows of $A$ are kept on the same worker node, then the last reduce operation can be executed locally and without any shuffling. The resulting execution plan under these assumptions is equivalent to the RMM. If the system chooses the sort-merge join algorithm instead, then the columns of $A$ will be distributed across the worker nodes. Consequently, the last reduce step causes a mandatory repartitioning. Then, the resulting execution plan is equivalent to the CPMM. Even though Gilbert has only one dataflow plan specified to implement the matrix multiplication, the Flink system can choose internally between the RMM and CPMM strategy. The strategy is selected by the optimizer which bases its decision on the data size and the partitioning, if available. %Consequently, Flink frees the programmer from this decision.

>>>>

%!TEX root=paper.tex
\section{Gilbert Runtime}
\label{sec:gilbertRuntime}

The Gilbert runtime is responsible for executing the compiled MATLAB code on a particular platform. For this purpose, it receives the intermediate representation of the code and translates it into the execution engine's specific format. Currently, Gilbert supports three different execution engines: \emph{ReferenceExecutor}, \emph{FlinkExecutor} and \emph{SparkExecutor}. The \emph{ReferenceExecutor} executes the compiled MATLAB code locally. For the distributed execution Gilbert supports Apache Spark and Apache Flink as backends. The \emph{ReferenceExecutor} is an interpreter for the intermediate code.
It takes the dependency tree of a MATLAB program and executes it by evaluating the operators bottom-up. In order to evaluate an operator, the system first evaluates all inputs of an operator and then executes the actual operator logic. Since the program is executed locally, the complete data of each operator is always accessible and, thus, no communication logic is required. 
All input and output files are directly written to the local hard drive. 
In contrast to the \emph{ReferenceExecutor}, the \emph{FlinkExecutor} and \emph{SparkExecutor} execute the program in parallel. 
It takes the dependency tree of a MATLAB program and translates it into a Flink and Spark program, respectively. 
After the program is generated, it is issued to the respective system for parallel execution. 
Consequently, data structures are needed which can be distributed across several nodes and represent the commonly used linear algebra abstractions, such as vectors and matrices. 
Moreover, the linear algebra operations have to be adjusted so that they keep working in a distributed environment. 
Fortunately, Flink and Spark offer a rich and expressive API to easily realize distributed computations.
Since both systems offer a similar set of programming primitives, they can operate on the same data structures. 
Furthermore, most of the linear algebra operations are implemented in a similar fashion.
The details of the distributed data structures and operations are explained in \cref{sec:DistributedMatrixRepresentation} and \cref{sec:LinearAlgebraOperations}.

@@

One aspect of writing distributed algorithms is how the relevant data is distributed across several worker nodes. Since the distribution pattern directly influences the algorithms, one cannot consider them independently from one another. In Gilbert's use case, the main data structure are matrices. Thus, a partitioning for matrices has to be conceived which enables efficient algorithms working on distributed data. Looking at a matrix, one easily finds a multitude of different partition schemes. A first idea could be to partition a matrix according to their rows or columns, as it is depicted in \cref{fig:rowPartitioning} and \cref{fig:columnPartitioning}. This scheme allows to represent a matrix as a set of vectors which is stored in a distributed fashion.
Furthermore, it allows an efficient realization of cellwise operations, such as $+,-,/$ or $.*$. In order to calculate the result of such an operation, we only have to join the corresponding rows of both operands and execute the operation locally for each pair of rows.

@@

This approach, however, unveils its drawbacks when multiplying two equally partitioned matrices $A$ and $B$. In such a case, the row $r$ of $A$ and the complete matrix $B$ is needed to calculate the resulting row with index $r$. This circumstance implies a complete repartitioning of $B$. The repartitioning is especially grave, since $B$ has to be broadcasted to every row of $A$. In order to quantify the repartitioning costs of such a matrix multiplication, a simple cost model is developed. First of all, it is limited to modeling the communication costs, since network I/O usually constitutes the bottleneck of distributed systems and is thus the dominating factor. For the sake of simplicity, the multiplication of two quadratic matrices $A \in \mathbb{R}^{n\times n}$ and $B \in \mathbb{R}^{n\times n}$ giving the matrix $C\in \mathbb{R}^{n \times n}$ is considered. Moreover, we assume that there are $n$ worker nodes available, each of which receiving a single row of $A$ and $B$, respectively. Thus, $A$ and $B$ are row-wise partitioned. We further assume that rows with the same index are kept on the same worker node. Each row $a_r$ requires the complete knowledge of matrix $B$ in order to produce the row $c_r$. Therefore, every row $b_r$ has to be sent to all other worker nodes. Thus, the number of rows sent by each worker node is $n-1$. All of the $n$ worker nodes have to do the same. Consequently, the total number of sent messages is $n(n-1)$ and each message has a size of $n\alpha$ where $\alpha$ is the size of a matrix entry. Usually, each sending operation causes some constant overhead inflicted by resource allocation. Before sending the actual data over the network, memory to transfer the data to the network interface has to be allocated, network resources have to be reserved and a network connection with the remote peer has to be established. This overhead is denoted by $\Delta$. Since all sending operations occur in parallel, the costs caused by constant overhead are $(n-1)\Delta$. The total amount of data, which has to sent over the network, is $n^2(n-1)\alpha$. The network interconnection is assumed to guarantee every node a bandwidth $\nu$. Therefore, the time needed for sending the data is $\frac{n^2(n-1)\alpha}{\nu}$. These considerations lead to the following repartitioning cost model:

@@

Row and column partitioning are extreme forms of blocking. A less extreme form would be to split the matrix into equally sized quadratic blocks as shown in \cref{fig:quadraticBlockPartitioning}. In order to identify the individual blocks, each of them will be a block row and block column index assigned. Thus, blocking adds some memory overhead in the form of index information. The blocks are distributed across the worker nodes. The block size directly controls the granularity of the partitioning. Increasing the block size will reduce the memory overhead of distributed matrices while reducing the degree of parallelism. Thus, the user has to adjust the block size value depending on the matrix sizes and the number of available worker nodes in order to obtain best performance. The quadratic block partitioning has similar properties like the row- and column-wise partitioning scheme when it comes to cellwise operations. We simply have to join corresponding blocks with respect to the pair of block row and column index and execute the operation on matching blocks locally. But how does this pattern performs for matrix multiplications? The assumptions are the same as before and additionally it is assumed that $n$ is a square number. Since the matrices $A$, $B$ and $C$ are equally partitioned into square blocks, indices will henceforth reference the block and not the matrix entry. In order to calculate the block $c_{ij}$, we have to know the block row $a_{i}$ and the block column $b_{j}$. The resulting block will be stored on the node $n_{ij}$ which already contains the blocks $a_{ij}$ and $b_{ij}$. Thus, each node $n_{ij}$ has to receive the missing $2\left(\sqrt{n}-1\right)$ blocks from the block row $a_{i}$ and block column $b_{j}$. In total, all worker nodes have to send $2n\left(\sqrt{n}-1\right)$ blocks. Each block has the size $n\alpha$. The total communication costs comprises the transmission costs and the network overhead:

@@

In contrast to RMM, CPMM calculates the outer products between $A^k$ and $B_k$ for all $k$. A mapper can group the $A^k$ and $B_k$ together so that a reducer can compute the outer products. Consequently, this method does not replicate any data. The outer product produces intermediate result matrices $C^k$ which have to be added up to produce the final result $C=\sum_{k=1}^{l} C^k$. This summation can be achieved by a subsequent reducer. The whole process is illustrated in \cref{fig:CPMM}. The methods RMM and CPMM differ in terms of network communication. The former method can be realized within a single MapReduce job whereas the latter requires two. Neither RMM nor CPMM is always superior. The optimal matrix multiplication strategy depends on the matrix size of its operands $A$ and $B$. Fortunately, Flink and Spark exhibit a little bit more flexibility in terms of higher order functions. Looking at the definition of the matrix multiplication for $C_{ij}=\sum_{k=1}^{l}A_{ik}\times B_{kj}$, it can be seen that every $A_{ik}$ has to be joined with its corresponding $B_{kj},\forall k$. This pairwise mapping can be easily achieved by using the join function. The join-key is the column index of $A$ and the row index of $B$. The joiner calculates for each matching pair $A_{ik}$ and $B_{kj}$ an intermediate result $C_{ij}^k$. Grouping the intermediate results with respect to the index pair $(i,j)$ allows us to compute the final result in a subsequent reduce step. The overall algorithm strongly resembles the CPMM.

Flink supports several execution strategies for the higher-order functions. 
A cost-based optimizer selects the best strategies prior to execution. 
One possible optimization concerns the join function. 
The join can either be realized using a broadcast hash join, a repartitioning hash join or a repartitioning sort-merge join algorithm depending on the current partitioning and the input data sizes. 
If one input data is relatively small compared to the other input, it is usually more efficient to use the broadcast hash join algorithm. 
Without loss of generality, we assume that the matrix $B$ constitutes such a small input. 
If we further assume that the block rows of $A$ are kept on the same worker node, then the last reduce operation can be executed locally and without any shuffling. 
The resulting execution plan under these assumptions is equivalent to the RMM. 
If the system chooses any of the repartitioning join algorithm instead, then the columns of $A$ will be distributed across the worker nodes. 
Consequently, the last reduce step causes a mandatory repartitioning. 
Then, the resulting execution plan is equivalent to the CPMM. 
Even though Gilbert has only one dataflow plan specified to implement the matrix multiplication, the Flink system can choose internally between the RMM and CPMM strategy. 
The strategy is selected by the optimizer which bases its decision on the data size and the partitioning, if available. %Consequently, Flink frees the programmer from this decision.
